Combining checkpointing and data compression\\ for large scale seismic inversion

### 2018-10-05 paper/paper.tex
fixed reference
Finishing up

Another way of using compression to speed up seismic inversion or other adjoint-based methods is to use it \emph{instead} of checkpointing. If the compression ratio is sufficient to fit the entire data in memory, checkpointing is no longer necessary. Previous work has discussed this situation in the context of computational fluid dynamics~\cite{cyr2015towards} and seismic inversion using compression method specifically designed for
wavefields~\cite{dalmau2014lossy,boehm2016wavefield}.

@@

We present test cases that use a TODO describe Richter
and TODO describe JLSE/SKX node system, henceforth called Richter and Skylake,
respectively. We discuss performance results for varying error tolerances for
the lossy ZFP compression algorithm ranging from TODO to TODO, and different
types of forward and adjoint computations that vary in their ratio between
compute cost and state size.

@@

\section{Test case using Devito and pyRevolve}
We use Devito~\cite{devito-api,devito-compiler} to solve forward and adjoint
wave equation problems. Devito is a domain-specific language that enables the
rapid development of finite-difference solvers from a high-level description
of partial differential equations. The simplest version of the seismic wave
equation is the acoustic isotropic wave equation defined as:
\begin{equation}
m(x)\frac{\partial^2 u(t, x)}{\partial t^2} - \Laplace u(t, x) = q(t, x),
\label{eqn:wave}
\end{equation}
where $m(x) = \frac{1}{c^2(x)}$ is the squared slowness, $c(x)$ the spatially
dependent speed of sound, $u(t, x)$ is the pressure wavefield, $\Laplace u(t, x)$ denotes the laplacian of the wavefield and $q(t, x)$ is a source term.
Some of our experiments in Section TODO are performed using a more accurate
and more complex version of this equation called Tilted Transverse Isotropy
(TTI)~\cite{zhang2011stable} that takes into account the anisotopic propagation of waves in the earth subsurface (directional dependency of the speed of sound). We leave the TTI equations out of this paper
for brevity.

@@

The highest compression efficiency TODO define this is achieved in the
fixed-tolerance mode albeit with highly unpredictable compression ratios. Figure
\ref{fig:decompressed_error} shows the spatial distribution of the error after
compression and decompression, compared to the original field, using
fixed-tolerance mode.

In our experiments using Devito, the time required to compress a checkpoint was
in the same orderof magnitude as the time taken to compute a timestep.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\linewidth]{images/tolerance-cf-richter.pdf}
\end{center}
\caption{Compression ratios achieved on compressing the wavefield. The
highlighted point corresponds to the setting used for the other results here
unless otherwise specified.}
\label{fig:tolerance_cf_plot}
\end{figure}

@@

In an ideal implementation, the storage overhead $\mathbf{O}_S$ might be zero, since the computation could
be done ``in-place'', but in practice, checkpoints are generally stored in a separate section of memory and they
need to be transferred to a ``computational'' section of the memory where the compute is performed, and then
the results copied back to the checkpointing memory. This copying is a common feature of checkpointing
implementations, and might pose a non-trivial overhead in scenarios where the computation is heavily bandwidth-bound. 
This storage overhead is given by
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot \frac{\mathbf{S}}{\mathbf{B}} +
\mathbf{N}_R \cdot \frac{\mathbf{S}}{\mathbf{B}}.
\label{eqn:storage}
\end{equation}
where $\mathbf{N}_W$ is the total number of times Revolve writes checkpoints for a single run, $ \mathbf{N}_R$ 
is the number of times checkpoints are read, and $\mathbf{B}$ is the memory bandwidth of the target system.

The total time to solution in this scenario becomes
\begin{equation}
T_R = 2 \cdot \mathbf{C} \cdot \mathbf{N} + \mathbf{O}_R(N, M) + \mathbf{O}_S.
\end{equation}

By using compression, the size of each checkpoint is reduced and therefore the
number of checkpoints available is increased ($M$ in equation
\ref{eqn:recompute}). This reduces the recomputation overhead $\mathbf{O}_R$,
while at the same time adding overheads related to compression and decompression
in $\mathbf{O}_S$.
To be beneficial, the reduction in $\mathbf{O}_R$ must offset the increase in 
$\mathbf{O}_S$, leading to an overall decrease in the time to solution $T$.

@@

\section{Results}
We can distinguish three different scenarios, depending on the amount of
available memory.
\begin{enumerate}
\item If the memory is insufficient even with compression to store the entire
trajectory, one can either use checkpointing only, or combine checkpointing with
compression.
\item If the available memory is not sufficient to store the uncompressed
trajectory, but large enough to store the entire compressed trajectory, we study
the two possible strategies: Either use compression only, or use checkpointing
only.
\item If the available system memory is large enough to hold the entire
uncompressed trajectory, neither compression nor checkpointing is necessary.
\end{enumerate}
All three scenarios can be seen in figure~\ref{fig:varying_memory}.
The second scenario was studied in previous work~\cite{cyr2015towards}, while
the combined method is also applicable to the first scenario, for which previous
work has only used checkpointing without compression.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{images/varying-memory.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  memory. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. The different curves represent kernels with differing
compute times (represented here as a factor of the sum of compression
and decompression times). The first vertical line at $~$ 53GB marks the
spot where the compressed wavefield can completely fit in memory and
Revolve is unnecessary if using compression. The second vertical line
at $~$ 2.2 TB marks the spot where the entire uncompressed wavefield can
fit in memory and neither Revolve nor compression is necessary. The
region to the right is where these optimisations are not necessary or
relevant. The middle region has been the subject of past studies using
compression in adjoint problems. The region to the left is the focus
of this paper.}
\label{fig:varying_memory}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{images/varying-compute.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  compute cost. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. The benefits of compression drop rapidly if the
computational cost of the kernel that generated the data is much lower
than the cost of compressing the data. For increasing computational
costs, the benefits are bounded.}
\label{fig:varying_compute}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{images/varying-nt.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  number of timesteps to be reversed. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. It can be seen that compression becomes more
beneficial as the number of timesteps is increased.}
\label{fig:varying_nt}
\end{figure}

\section{Conclusions and Future work}
We use lossy compression to reduce the computational overhead of checkpointing
in an adjoint computation used in full waveform inversion, a common method in
seismic imaging applications whose memory footprint commonly exceeds the
available memory size in high performance computing systems. We also developed a
performance model that computes whether or not the combination of compression
and checkpointing will outperform pure checkpointing or pure compression in a
variety of scenarios, depending on the available memory size, computational
intensity of the application, and compression ratio and throughput of the
compression algorithm.

Our current result has several limitations that we plan to address in future
work:
\begin{itemize}
\item We do not discuss the accuracy of the results after decompression. This
depends on the application, compression algorithm, and affects the achievable
compression ratios. Our performance model only requires knowledge of the
compression time and ratio, and it is up to the user of this model to determine
what accuracy is needed and thus what compression ratio is realistic for their
application.
\item We only present results using the ZFP compression library. Again, the
performance model is independent of this, but it would be interesting to try a
greater variety of compression algorithms.
\item ZFP currently only supports serial decompression. If ZFP supported
parallel decompression, our experiments would likely show a geater region in
which the combined method is faster than pure checkpointing. Furthermore, ZFP
only supports fields with up to three dimensions, and it is possible that
exploiting similarities between fields at different time steps would yield a
better compression ratio.
\item Our performance model is based on uniform compression ratios and times.
However, many applications, including FWI, are likely to have initial conditions
that contain little information and are easily compressed, and the compression
ratio gradually declines as the field becomes more complex over time. We have
based our experiments on the final wave field, which is presumably the most
difficult to compress.
\item In comparing pure compression with pure checkpointing, we assume that
every checkpoint would be compressed and decompressed. However, in scenarios
where the available memory is only slightly less than the required memory, an
implementation that compresses only a subset of the checkpoints might outperform
the expectations of our model. 
\item We do not discuss multi-level checkpointing, where some checkpoints are
stored on a slower, larger device such as a solid state drive. We expect
compression to be particularly beneficial in these scenarios due to the reduced
data transfer sizes.
\end{itemize}

>>>>

Another way of using compression to speed up adjoint-based methods is to use it \emph{instead} of
checkpointing. If the compression ratio is sufficient to fit the
entire data in memory, checkpointing is no longer necessary. Previous
work has discussed this in the context of computational
fluid dynamics~\cite{cyr2015towards} and seismic inversion using
compression algorithms specifically designed for
wavefields~\cite{dalmau2014lossy,boehm2016wavefield}.

@@

For benchmarking we used a dual-socket
Intel(R) Xeon(R) Platinum 8180M @ 2.50 Ghz (28 cores each), henceforth
called Skylake. We discuss performance results for varying error tolerances for
the lossy ZFP compression algorithm ranging from $10^0$ to $10^{-15}$, and different
types of forward and adjoint computations that vary in their ratio between
compute cost and state size.

@@

\section{Test case using Devito and pyRevolve}
We use Devito~\cite{devito-api,devito-compiler} to solve forward and adjoint
wave equation problems. Devito is a domain-specific language that enables the
rapid development of finite-difference solvers from a high-level description
of partial differential equations. The simplest version of the seismic wave
equation is the acoustic isotropic wave equation defined as:
\begin{equation}
m(x)\frac{\partial^2 u(t, x)}{\partial t^2} - \Laplace u(t, x) = q(t, x),
\label{eqn:wave}
\end{equation}
where $m(x) = \frac{1}{c^2(x)}$ is the squared slowness, $c(x)$ the spatially
dependent speed of sound, $u(t, x)$ is the pressure wavefield, $\Laplace u(t, x)$ denotes the laplacian of the wavefield and $q(t, x)$ is a source term.
Some of our experiments in Section \ref{sec:results} are performed using a more accurate
and more complex version of this equation called Tilted Transverse Isotropy
(TTI)~\cite{zhang2011stable} that takes into account the anisotopic propagation of waves in the earth subsurface (directional dependency of the speed of sound). We leave the TTI equations out of this paper
for brevity.

@@

The compression ratio achieved per unit time spent during compressing
and decompressing was seen to be highest in the fixed-tolerance mode
albeit with highly unpredictable compression ratios. Figure
\ref{fig:decompressed_error} shows the spatial distribution of the
error after compression and decompression, compared to the original
field, using fixed-tolerance mode.

In our experiments using Devito, the time required to compress a checkpoint was
in the same order of magnitude as the time taken to compute a timestep.

\begin{figure}
\begin{center}
\includegraphics[width=0.9\linewidth]{images/tolerance-cf-richter.pdf}
\end{center}
\caption{Compression ratios achieved on compressing the wavefield. We
  define compression ratio as the ratio between the size of the
  uncompressed data and the compressed data. The dashed line
  represents no compression. The highlighted point corresponds to the
  setting used for the other results here unless otherwise specified.}
\label{fig:tolerance_cf_plot}
\end{figure}

@@

In an ideal implementation, the storage overhead $\mathbf{O}_S$ might be zero, since the computation could
be done ``in-place'', but in practice, checkpoints are generally stored in a separate section of memory and they
need to be transferred to a ``computational'' section of the memory where the compute is performed, and then
the results copied back to the checkpointing memory. This copying is a common feature of checkpointing
implementations, and might pose a non-trivial overhead in scenarios where the computation is heavily bandwidth-bound. 
This storage overhead is given by
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot \frac{\mathbf{S}}{\mathbf{B}} +
\mathbf{N}_R \cdot \frac{\mathbf{S}}{\mathbf{B}}.
\label{eqn:storage}
\end{equation}
where $\mathbf{N}_W$ is the total number of times Revolve writes
checkpoints for a single run, $ \mathbf{N}_R$ is the number of times
checkpoints are read, and $\mathbf{B}$ is the memory bandwidth of the
target system. The total time to solution becomes
\begin{equation}
T_R = 2 \cdot \mathbf{C} \cdot \mathbf{N} + \mathbf{O}_R(N, M) + \mathbf{O}_S.
\end{equation} 
By using compression, the size of each checkpoint is reduced and therefore the
number of checkpoints available is increased ($M$ in equation
\ref{eqn:recompute}). This reduces the recomputation overhead $\mathbf{O}_R$,
while at the same time adding overheads related to compression and decompression
in $\mathbf{O}_S$.
To be beneficial, the reduction in $\mathbf{O}_R$ must offset the increase in 
$\mathbf{O}_S$, leading to an overall decrease in the time to solution $T$.

@@

\section{Results}
\label{sec:results}
We can distinguish three different scenarios, depending on the amount of
available memory.
\begin{enumerate}
\item If the memory is insufficient even with compression to store the entire
trajectory, one can either use checkpointing only, or combine checkpointing with
compression.
\item If the available memory is not sufficient to store the uncompressed
trajectory, but large enough to store the entire compressed trajectory, we study
the two possible strategies: Either use compression only, or use checkpointing
only.
\item If the available system memory is large enough to hold the entire
uncompressed trajectory, neither compression nor checkpointing is necessary.
\end{enumerate}
All three scenarios can be seen in figure~\ref{fig:varying_memory}.
The second scenario was studied in previous work~\cite{cyr2015towards}, while
the combined method is also applicable to the first scenario, for which previous
work has only used checkpointing without compression.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.9\linewidth]{images/varying-memory.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  memory. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. The different curves represent kernels with differing
compute times (represented here as a factor of the sum of compression
and decompression times). The first vertical line at $~$ 53GB marks the
spot where the compressed wavefield can completely fit in memory and
Revolve is unnecessary if using compression. The second vertical line
at $~$ 2.2 TB marks the spot where the entire uncompressed wavefield can
fit in memory and neither Revolve nor compression is necessary. The
region to the right is where these optimisations are not necessary or
relevant. The middle region has been the subject of past studies using
compression in adjoint problems. The region to the left is the focus
of this paper.}
\label{fig:varying_memory}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\linewidth]{images/varying-compute.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  compute cost. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. The benefits of compression drop rapidly if the
computational cost of the kernel that generated the data is much lower
than the cost of compressing the data. For increasing computational
costs, the benefits are bounded.}
\label{fig:varying_compute}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=0.9\linewidth]{images/varying-nt.pdf}
\end{center}
\caption{The speedups predicted by the performance model for varying
  number of timesteps to be reversed. The baseline
(1.0) is the performance of a Revolve-only implementation under the
same conditions. It can be seen that compression becomes more
beneficial as the number of timesteps is increased.}
\label{fig:varying_nt}
\end{figure}

\section{Conclusions and Future work}
We use lossy compression to reduce the computational overhead of
checkpointing in an adjoint computation used in full waveform
inversion, a common method in seismic imaging applications whose
memory footprint commonly exceeds the available memory size in high
performance computing systems. We also developed a performance model
that computes whether or not the combination of compression and
checkpointing will outperform pure checkpointing or pure compression
in a variety of scenarios, depending on the available memory size,
computational intensity of the application, and compression ratio and
throughput of the compression algorithm. Our current result has
several limitations that we plan to address in future work:
\begin{itemize}
\item We do not discuss the accuracy of the results after decompression. This
depends on the application, compression algorithm, and affects the achievable
compression ratios. Our performance model only requires knowledge of the
compression time and ratio, and it is up to the user of this model to determine
what accuracy is needed and thus what compression ratio is realistic for their
application.
\item We only present results using the ZFP compression library. It
would be interesting to try a greater variety of compression
algorithms.
\item ZFP only supports serial decompression. If ZFP supported
parallel decompression, our experiments would likely show a geater region in
which the combined method is faster than pure checkpointing. Furthermore, ZFP
only supports fields with up to three dimensions, while
exploiting similarities between fields at different time steps may yield a
better compression ratio.
\item Our performance model is based on uniform compression ratios and
times.  However, many applications, including FWI, are likely to have
initial conditions that contain little information and are easily
compressed, and the compression ratio gradually declines as the field
becomes more complex. We based our experiments on the final wave
field, which is presumably difficult to compress.
\item In comparing pure compression with pure checkpointing, we assume
that every checkpoint is compressed and decompressed. However, if the
available memory is only slightly less than the required memory, an
implementation that compresses only a subset of the checkpoints might
outperform the expectations of our model.
\item We do not discuss multi-level checkpointing, where some
checkpoints are stored on a slower, larger device. We expect
compression to be beneficial in these scenarios due to reduced data
transfer sizes.
\end{itemize}
